Gringo 2.0

### 2008-08-04 doc/documentation/paper.tex
further improvements, spellcheck again, removed Benchmarks, etc... all doc
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@417 3e02f596-3131-0410-8352-e04eeaec37cc

\begin{document}
\maketitle
\tableofcontents
\section{Introduction}
Answer Set Programming (ASP) \cite{baral02a} is a declarative approach for modeling and solving search problems, represented as logic programs.
An ASP system usually consists of a grounder and a solver.
First, a grounder translates a non-ground problem description into a propositional program, which can be processed by the second component of the system, the solver. This is illustrated in figure (\ref{fig:ASP}).
\begin{figure}[h]
\[
\xymatrix{*+[F]\txt{Problem\\representation}\ar[rr]^{Grounder}&&
		*+[F]\txt{Intermediate\\ representation} \ar[rr]^{Solver}&&
		*+[F]\txt{Variable\\ assignment}}
\]
\caption{Architecture of an ASP System}
\label{fig:ASP}
\end{figure}
gringo2.0 is a state of the art grounder and has many features that makes writing logic programs easy.
It was developed by Roland Kaminski and is orientated on the ideas of gringo1.0.
The goal of this work is to give an overview of the features supported by gringo2.0. A broad set of potentialities for writing encodings for logic programs exists. Currently, no other grounder supports such a rich featured language.
gringo2.0 supports several output formats and even a human readable format for debugging purposes.
It also has an internal interface to the solver clasp \cite{gekanesc07b}, so the output can directly by parsed to the solver and answer sets can be computed.
Furthermore gringo2.0 can be used with clasp incrementally. Detailed information about incremental solving can be found here \cite{IncrementalASP} 
First, the general restrictions to the input language are stated. Then a brief description of how to use gringo2.0 is given.
In the next chapter the syntax of the supported language is described in detail and a short introduction of how to write a logic program is given. The paper closes with some examples of logic programs and gives an outlook of future work that has to be done in the area of grounding logic programs.
The syntax of the gringo2.0 input language can be found in the appendix \ref{appendix}.

\section{$\lambda$-Restriction}
\label{domainrestricted}
There exist two major grounders for ASP besides gringo1.0, lParse \cite{lparseManual} and  the DLV grounding module\cite{DLV}.
To ensure the existence of an equivivalent ground instantiation of a program both grounders except a subclass of logic programs. For lParse the logic programs must be $\omega$-restricted\cite{omega} while DLV can handle safe programs\cite{safe}.
gringo2.0 verifies that the input program is $\lambda$-
restricted \cite{gringopaper}, so that the existence of a finite equivalent ground instantiation is guaranteed. 
$\lambda$-restrictedness extends the capabilities of $\omega$-restricted programs. The definition of $\lambda$-restrictedness can be found in \cite{gringopaper} and was expanded for aggregates and disjunctions.
$\lambda$-restrictedness can not be compared with the safe programs of DLV since there exist programs that are safe and not $\lambda$-restricted and vice versa.

\section{Using gringo2.0}
\subsection{Command line Interface}
gringo2.0 can be started on the command line and has some parameters that can be useful.
The first parameters are gringo2.0 specific parameters and consist of a number of options and any arbritary number of filenames of the logic program. For the special purpose that gringo2.0 is used with its clasp module a double minus ``\texttt{--}'' is used to seperate the clasp parameters from the gringo2.0 parameters.
If no filename is given, gringo2.0 reads from standard input.
\begin{verbatim}
 ./gringo (gringo option|file)* [--[ clasp options]]
\end{verbatim}
\subsection{Options}

@@

\paragraph{\texttt{--convert}}
If this parameter is used, the grounder expects an already grounded program as input.
For converting an already grounded logic program into an output format supported by gringo2.0 this is the right option. This mode saves a lot of memory but does not ground the program.
\paragraph{\texttt{-l}}
This mode prints smodels conform output. For a definition of the output format see \cite{lparseManual}.
Since some features of gringo2.0 are not natively supported for this output language, new atoms are introduced to create an equivalent program.\newline
This is the current default output format of gringo2.0.
Sum aggregates with negative weights are handled like weight constraint in lparse for this output format. So a transformation as explaned in \cite{lparseManual} is used to get rid of the negative weights.
\paragraph{\texttt{-c}}
In this mode, the output is directly parsed to clasp, an ASP solver. For solving logic programs, no intermediate files are produced and the answer sets are printed directly.
With this option enabled, all parameters for clasp are valid and can be declared in the command line. Therefore a double minus ``\texttt{--}'' must be given after all gringo2.0 related parameters are set. After them the clasp parameters follow.
For example:
\begin{verbatim}
 gringo -c --const n=5 -- --supp-models 3
\end{verbatim}
Furthermore all restriction on smodels output hold as clasps internal interface currently reads smodels format.
\paragraph{\texttt{-p}}
Parameter ``\texttt{-p}'' invokes the plain output method, so the output is printed in human readable format, which is also the input format for gringo2.0. The output is variable free and some simplifications on the program were made. To convert the plain format into a format readable by a solver use the \texttt{--convert} option.
\paragraph{\texttt{-g [1-7]}}
This parameter invokes the new output format for the intermediate languages ASPils \cite{atleastthisreferenceishouldgive}. The optional number can range from $1$ to $7$ and refers to one of the following normal forms (Simple=1, SModels=2, DisjunctiveLogicProgram=3, CModels=4, CModelsExtended=5, DLV=6, Conglomeration=7). If no normal form is specified, conglomeration(7)  is assumed.

@@

\begin{itemize}
 \item 
\begin{verbatim}
 "Warning: sum with negative bounds in the body of a rule"
\end{verbatim}
This warning occurs if a sum aggregate with negative weights has been found and the output format smodels is used or the output is redirected to clasp. In this format negative bounds are compiled away using the transformation described in \cite{lparseManual}. This may lead to counterintuitive results.
 \item 
\begin{verbatim}
 "Warning: multiple definitions of #const n"
\end{verbatim}
This warning is printed if more than one \#const statement is used to set the same constant.
The last attached value is used for grounding.
\item
\begin{verbatim}
 "Warning: p/3 is never defined"
\end{verbatim}
This warnings states that an atom is used in the body of a rule that never occurs in the head.
The predicate literal is given in the warning. The predicate will be removed trhough optimization.
\item
\begin{verbatim}
 "got ground program i hope u have enough memory :)"
\end{verbatim}
Use the command line option ``--convert'' to convert a ground program into another format.
For converting already ground programs to an output language a special module has been implemented that uses far less memory. Because trying to ground an already variable free program could need a huge amount of memory.

@@

\subsection{Errors}
gringo2.0 stops the execution and prints an error message to standard error output.
\begin{itemize}
 \item 
\begin{verbatim}
 "the following rule cannot not be grounded,
  non domain predicates : { ... }"
\end{verbatim}
This error message gives a list of predicates that are not domain restricted.
You explicitly have to give a domain for those predicates.
 \item 
\begin{verbatim}
 "the following rule cannot not be grounded,
  weakly restricted variables: { ... }"
\end{verbatim}
This error message gives a list of variables that are not lambda restricted. 
\item
\begin{verbatim}
 "syntax error on line Y column Z unexpected token: X"
\end{verbatim}
The parser can not read the input format. Some unexpected token occured in the input file.
\newline
\newline
The next errors all belong to the ASPils output format.
Some constructs in the input language can not be expressed in certain normal forms.
Another normal form as to be chosen to support this feature.
\begin{verbatim}
 "Disjunction not allowed on this normal form,
  please choose normal form 2 or 4-7."
 "Count aggregate/cardinality constraint not allowed
  in this normal form,
  please choose normal form 3-7."
 "Count aggregate/cardinality constraint not allowed 
  with non trivial upper bound in this normal form,
  please choose normal form 5-7."
 "Count aggregate/cardinality constraint not allowed 
  with negative lower bound in this normal form,
  please choose normal form 5-7."
 "Sum aggregate/weight constraint not allowed
  in this normal form,
  please choose normal form 3-7."
 "Sum aggregate/weight constraint not allowed 
  with non trivial upper bound in this normal form,
  please choose normal form 5-7."
 "Sum aggregate/weight constraint not allowed 
  with negative lower bound in this normal form,
  please choose normal form 5-7"
 "Sum aggregate/weight constraint not allowed
  with negative weights in this normal form,
  please choose normal form 5 or 7."
 "Max aggregate not allowed in this normal form,
  please choose normal form 6 or 7."
 "Max aggregate not allowed
  with negative weights in this normal form,
  please choose normal form 7."
 "Min aggregate not allowed in this normal form,
  please choose normal form 6 or 7."
 "Min aggregate not allowed 
  with negative weights in this normal form,
  please choose normal form 7."
 "Optimize statement not allowed in this normal form,
  please choose normal form 3-7."
\end{verbatim}
%\begin{verbatim}
% "Times aggregate not allowed in this normal form."
% "Times aggregate not allowed
%  with negative weights in this normal form."
%\end{verbatim}
\end{itemize}

@@

\section{Syntax}
The basic elements of a logic program and its syntax is now described.
Disjunctive logic programs are based on an alphabet \cite{syntax} consisting of:
\begin{itemize}
\item a set P of predicate symbols,
\item a set F of function symbols,
\item a set V of variables,
\item a set C of constants,
\item a set \{ ``$\mid$'', ``,'' ,$\leftarrow$, not \} of connectives, and
\item a set \{ ``('', ``)'', ``,'' \} of punctuation symbols.
\end{itemize}

TODO: Which greek letters for P, F, L and C
\paragraph{Predicates and function symbols}
Predicates and function symbols have a predicate symbol/name and an arity.
For the function symbol $f$ with arity 3 we write $f/3$.
The names of the function symbols and predicates have to start with either a lower case letter or an underscore and can contain digits and letters.
With $P$ we denote to the set of predicates and $F$ denotes the set of function symbols.
\paragraph{Constants}
Constants are function symbols with an arity of zero.
The name of a constant can also be encompassed by double quotation marks, which is sometimes helpful when using constants with spaces. It has to be remarked that a constant with quotation marks is never equal to a constant without, so ``\texttt{a}'' is not equal to \texttt{a}.
$C$ is the set of constants
\paragraph{Variables}
Variables have to start with a capitalized letter and can contain digits and letters.
The set of variables is denoted by $V$.
The language over $P$, $F$ and $C$ is denoted by $L(P,F,C)$.
\paragraph{Terms}
The building blocks of the language are terms.
Each constant $c \in C$ and each function symbol $f \in F$ is a term.
Furthermore $f(t_1, \dots, t_n)$ is a term if $f/n \in F$ and $t_i \mid i \in \{1 \dots n\}$ are terms.
Since a term is defined recursively, it can be nested.
A term with the first parameter being a constant $a$, the others being terms $f(b,X)$ and $g(f(X), Y))$ would be expressed this way.
\begin{verbatim}
 p(a,f(b,X), g(f(X),Y)).
\end{verbatim}
\paragraph{Atoms}
An atom over $L(P, F, C)$ is of the form $p(t_1, \dots, t_n)$ where $p/n$ is a predicate
symbol with arity $n$, and $t_i \mid i \in \{1 \dots n\}$ are terms.
\paragraph{Rules}
A rule over the language $L(P,F,C)$ has the form
$$
A_0 \mid \dots \mid A_l \leftarrow A_{l+1}, \dots, A_m, not\ A_{m+1}, \dots, not\ A_n.
$$
where $A_0$ to $A_n$ are atoms. \newline
A normal logic program $\Pi$ over $L(P,F,C)$ is a finite set of rules over $L(P,F,C)$.

@@

\subsection{How to write a logic program}
Writing a logic program to solve a problem is rather describing a problem than describing an algorithm to solve a problem.
Therefore the problem must be encoded as a set of rules.
Typically these are divided into a problem description and an instance of the problem.
The solution to a problem can then be read from the answer sets of the program.
The procedure of writing a logic program shall now be explained with a little example.
We want to write a problem encoding for the 3-colorability problem of graph coloring.
Given an arbitraty graph, no two connected edges may have the same color.
The most common technique is the ``generate \& test'' approach. Therefore, in the generator, rules are created that can derive a superset of the solution to the problem. In the tester all solutions that do not fit the constraints are omitted.
\newline
Usually you start with the non instance domains, in our example the colors we use.
\begin{verbatim}
 color(red). color(green). color(blue).
\end{verbatim}
After that, we start with the generator.
Therefore we have to color all the vertices, but we must give attention that each vertex has only one color.
We can do this by simply saying that a vertex is colored with one color $C$, and that the same vertex does not have any of the other colors $D$ or $E$.
\begin{verbatim}
 colored(V,C) :- not colored(V,D), not colored(V,E),
                 C != D, C != E, D != E,
                 color(E), color(C), color(D), vertex(V).
\end{verbatim}
As you can see, you have to give the domain of the variables in the rule.
The variable $V$ is used for vertices and the variables $C$, $D$ and $E$ are used for colors.
After every vertex now has a color we would get all possible solutions of all vertices colorfully painted.
We now need to restrict our solutions in the tester by saying that two vertices that are connected by an edge should not have the same color.
We can do this by using an integrity constraint.
An integrity constraint is a rule with an empty head, and if an assignment of variables fulfills the body, the assignment is not a solution of our problem.
\begin{verbatim}
 :- edge(V,U), colored(V,C), colored(U,C),
    vertex(V), vertex(U), color(C).
\end{verbatim}
Here we have the edge between two vertices $V$ and $U$, and if both are colored with the same color $C$ the assignment is not a solution to our problem.
\newline
We have finally written our problem description, all we need now is a problem instance. This step is fairly simple.
We just need to define several vertices and edges that do connect the vertices.
For the following simple graph
\[
 \xymatrix{1\ar@{-}[dr]\ar@{-}[d]\ar@{-}[r]&2\ar@{-}[dl]\\
           3&4}
\]

@@

\subsection{Features}
To ease the process writing logic programs some features in the input syntax are introduced and are now explained in detail.
For the most problems it is possible to write a much shorter version of the program with the extended syntax.
%Even a speedup in solving the instances of the problem can happen TODO ?
For a full syntax description in EBNF see appendix (\ref{appendix}).
\paragraph{Function Symbols}
Function symbols are fully supported with gringo2.0.
They can be finitly nested and also be used as domain predicates.
While in lParse function symbols are allowed in a very restricted way, no restrictions apply in gringo2.0.
The following program can not be grounded with lParse since $X$ is an unrestricted variable. With gringo2.0 the program is now groundable.
\begin{verbatim}
a(X) :- f(1,g(X)).
f(1,X) :- f(X).
f(g(2)).
\end{verbatim}
From the second rule the domain \texttt{f(1,g(2))} is generated.
This domain can be used for the first rule and results in the grounded rule
\begin{verbatim}
 a(2) :- f(1,g(2)).
\end{verbatim}

@@

\paragraph{Aggregates}
An aggregate is a set or a multiset of atoms and has a lower and an upper bound.
The general syntax is
\begin{verbatim}
 lower aggregate_name [atom1=w1, ..., atomN=wN] upper
\end{verbatim}
If squared brackets are used, the atom list is interpretated as a set, otherwise, for curly brackets, as a multiset of atoms.
The lower and upper bound is optional, and can be omitted.

@@

\subparagraph{Min}
The ``min'' aggregate is determined by the minimum of all weights of atoms that evaluate to true.
For the minimum aggregate a multiset with weighted atoms is used. It has no shorthand syntax.
\subparagraph{Max}
The ``max'' aggregate is similiar to the minimum aggregate but considers the maximum of the weights.

@@

% fail if we havnt found a better solution
 :- costs(C), C [ altserves(R,Dist) : dist(Dist) : restaurant(R,RK) = Dist ].

@@

\section{Benchmarks}
The benchmarks were run on an Athlon XP 1700+ with 1GB of RAM.
TODO: Official (non internal wiki) link to the instances.

\input{benchmark2.result}
\section{Future Work}
$\Pi$-restrictedness
bottom up grounding
get rid of backward compatibility
more aggregates

\newpage
\textbf{\appendixname}
\begin{appendix}
\section{EBNF Syntax}
\label{appendix}
\input{appendix}
\end{appendix}

>>>>

\begin{document}
\maketitle
\tableofcontents
\section{Introduction}
Answer Set Programming (ASP) \cite{baral02a} is a declarative approach for modeling and solving search problems, represented as logic programs.
An ASP system usually consists of a grounder and a solver.
First, a grounder translates a non-ground problem description into a propositional program, which can be processed by the second component of the system, the solver. This is illustrated in figure (\ref{fig:ASP}).
\begin{figure}[h]
\[
\xymatrix{*+[F]\txt{Problem\\representation}\ar[rr]^{Grounder}&&
		*+[F]\txt{Intermediate\\ representation} \ar[rr]^{Solver}&&
		*+[F]\txt{Variable\\ assignment}}
\]
\caption{Architecture of an ASP System}
\label{fig:ASP}
\end{figure}
gringo2.0 is a state of the art grounder and has many features that makes writing logic programs easy.
It was developed by Roland Kaminski and is orientated on the ideas of gringo1.0.
Therefore it supports Bindersplitting, Backjumping and all other features already implemented in gringo1.0.\cite{gringopaper}
The goal of this work is to give an overview of the features supported by gringo2.0. A broad set of potentialities for writing encodings for logic programs exists. Currently, no other grounder supports such a rich featured language.
gringo2.0 supports several output formats and even a human readable format for debugging purposes.
It also has an internal interface to the solver clasp \cite{gekanesc07b}, so the output can directly by parsed to the solver and answer sets can be computed.
Furthermore gringo2.0 can be used with clasp incrementally. Detailed information about incremental solving can be found here \cite{IncrementalASP} 
First, the general restrictions to the input language are stated. Then a brief description of how to use gringo2.0 is given.
In the next chapter the syntax of the supported language is described in detail and a short introduction of how to write a logic program is given. The paper closes with some examples of logic programs and gives an outlook of future work that has to be done in the area of grounding logic programs.
The syntax of the gringo2.0 input language can be found in the appendix \ref{appendix}.

\section{$\lambda$-Restriction}
\label{domainrestricted}
There exist two major grounders for ASP besides gringo1.0, lParse \cite{lparseManual} and  the DLV grounding module\cite{DLV}.
To ensure the existence of an equivalent ground instantiation of a program both grounders except a subclass of logic programs. For lParse the logic programs must be $\omega$-restricted\cite{omega} while DLV can handle safe programs\cite{safe}.
gringo2.0 verifies that the input program is $\lambda$-
restricted \cite{gringopaper}, so that the existence of a finite equivalent ground instantiation is guaranteed. 
$\lambda$-restrictedness extends the capabilities of $\omega$-restricted programs. The definition of $\lambda$-restrictedness can be found in \cite{gringopaper} and was expanded for aggregates and disjunctions.
$\lambda$-restrictedness can not be compared with the safe programs of DLV since there exist programs that are safe and not $\lambda$-restricted and vice versa.

\section{Using gringo2.0}
\subsection{Command line Interface}
gringo2.0 can be started on the command line and has some parameters that can be useful.
The first parameters are gringo2.0 specific parameters and consist of a number of options and any arbitrary number of filenames of the logic program. For the special purpose that gringo2.0 is used with its clasp module a double minus ``\texttt{--}'' is used to separate the clasp parameters from the gringo2.0 parameters.
If no filename is given, gringo2.0 reads from standard input.
\begin{verbatim}
 ./gringo (gringo option|file)* [--[ clasp options]]
\end{verbatim}
\subsection{Options}

@@

\paragraph{\texttt{--convert}}
If this parameter is used, the grounder expects an already grounded program as input.
For converting an already grounded logic program into an output format supported by gringo2.0 this is the right option. This mode saves a lot of memory but does not ground the program.
\paragraph{\texttt{-l}}
This mode prints smodels conform output. For a definition of the output format see \cite{lparseManual}.
Since some features of gringo2.0 are not natively supported for this output language, new atoms are introduced to create an equivalent program.\newline
This is the current default output format of gringo2.0.
Sum aggregates with negative weights are handled like weight constraint in lparse for this output format. So a transformation as explained in \cite{lparseManual} is used to get rid of the negative weights.
\paragraph{\texttt{-c}}
In this mode, the output is directly parsed to clasp, an ASP solver. For solving logic programs, no intermediate files are produced and the answer sets are printed directly.
With this option enabled, all parameters for clasp are valid and can be declared in the command line. Therefore a double minus ``\texttt{--}'' must be given after all gringo2.0 related parameters are set. After them the clasp parameters follow.
For example:
\begin{verbatim}
 gringo -c --const n=5 -- --supp-models 3
\end{verbatim}
Furthermore all restriction on smodels output hold as clasps internal interface currently reads smodels format.
\paragraph{\texttt{-p}}
Parameter ``\texttt{-p}'' invokes the plain output method, so the output is printed in human readable format, which is also the input format for gringo2.0. The output is variable free and some simplifications on the program were made. To convert the plain format into a format readable by a solver use the \texttt{--convert} option.
\paragraph{\texttt{-g [1-7]}}
This parameter invokes the new output format for the intermediate languages ASPils \cite{atleastthisreferenceishouldgive}. The optional number can range from $1$ to $7$ and refers to one of the following normal forms (Simple=1, SModels=2, DisjunctiveLogicProgram=3, CModels=4, CModelsExtended=5, DLV=6, Conglomeration=7). If no normal form is specified, conglomeration(7)  is assumed.

@@

\begin{itemize}
 \item 
\begin{verbatim}
 "Warning: sum with negative bounds in the body of a rule"
\end{verbatim}
This warning occurs if a sum aggregate with negative weights has been found and the output format smodels is used or the output is redirected to clasp. In this format negative bounds are compiled away using the transformation described in \cite{lparseManual}. This may lead to counterintuitive results.
 \item 
\begin{verbatim}
 "Warning: multiple definitions of #const n"
\end{verbatim}
This warning is printed if more than one \#const statement is used to set the same constant.
The last attached value is used for grounding.
\item
\begin{verbatim}
 "Warning: p/3 is never defined"
\end{verbatim}
This warnings states that an atom is used in the body of a rule that never occurs in the head.
The predicate literal is given in the warning. The predicate will be removed through optimization.
\item
\begin{verbatim}
 "got ground program i hope u have enough memory :)"
\end{verbatim}
Use the command line option ``--convert'' to convert a ground program into another format.
For converting already ground programs to an output language a special module has been implemented that uses far less memory. Because trying to ground an already variable free program could need a huge amount of memory.

@@

\subsection{Errors}
gringo2.0 stops the execution and prints an error message to standard error output.
\begin{itemize}
 \item 
\begin{verbatim}
 "the following rule cannot not be grounded,
  non domain predicates : { ... }"
\end{verbatim}
This error message gives a list of predicates that are not domain restricted.
You explicitly have to give a domain for those predicates.
 \item 
\begin{verbatim}
 "the following rule cannot not be grounded,
  weakly restricted variables: { ... }"
\end{verbatim}
This error message gives a list of variables that are not lambda restricted. 
\item
\begin{verbatim}
 "syntax error on line Y column Z unexpected token: X"
\end{verbatim}
The parser can not read the input format. Some unexpected token occurred in the input file.
\newline
\newline
The next errors all belong to the ASPils output format.
Some constructs in the input language can not be expressed in certain normal forms.
Another normal form as to be chosen to support this feature.
\begin{verbatim}
 "Disjunction not allowed on this normal form,
  please choose normal form 2 or 4-7."
 "Count aggregate/cardinality constraint not allowed
  in this normal form,
  please choose normal form 3-7."
 "Count aggregate/cardinality constraint not allowed 
  with non trivial upper bound in this normal form,
  please choose normal form 5-7."
 "Count aggregate/cardinality constraint not allowed 
  with negative lower bound in this normal form,
  please choose normal form 5-7."
 "Sum aggregate/weight constraint not allowed
  in this normal form,
  please choose normal form 3-7."
 "Sum aggregate/weight constraint not allowed 
  with non trivial upper bound in this normal form,
  please choose normal form 5-7."
 "Sum aggregate/weight constraint not allowed 
  with negative lower bound in this normal form,
  please choose normal form 5-7"
 "Sum aggregate/weight constraint not allowed
  with negative weights in this normal form,
  please choose normal form 5 or 7."
 "Max aggregate not allowed in this normal form,
  please choose normal form 6 or 7."
 "Max aggregate not allowed
  with negative weights in this normal form,
  please choose normal form 7."
 "Min aggregate not allowed in this normal form,
  please choose normal form 6 or 7."
 "Min aggregate not allowed 
  with negative weights in this normal form,
  please choose normal form 7."
 "Optimize statement not allowed in this normal form,
  please choose normal form 3-7."
\end{verbatim}
%\begin{verbatim}
% "Times aggregate not allowed in this normal form."
% "Times aggregate not allowed
%  with negative weights in this normal form."
%\end{verbatim}
\end{itemize}

@@

\section{Syntax}
The basic elements of a logic program and its syntax is now described.
Disjunctive logic programs are based on an alphabet \cite{syntax} consisting of:
\begin{itemize}
\item a set $\mathcal{P}$ of predicate symbols,
\item a set $\mathcal{F}$ of function symbols,
\item a set $\mathcal{V}$ of variables,
\item a set $\mathcal{C}$ of constants,
\item a set \{ ``$\mid$'', ``,'' ,$\leftarrow$, not \} of connectives, and
\item a set \{ ``('', ``)'', ``,'' \} of punctuation symbols.
\end{itemize}

\paragraph{Predicates and function symbols}
Predicates and function symbols have a predicate symbol/name and an arity.
For the function symbol $f$ with arity 3 we write $f/3$.
The names of the function symbols and predicates have to start with either a lower case letter or an underscore and can contain digits and letters.
With $\mathcal{P}$ we denote to the set of predicates and $\mathcal{F}$ denotes the set of function symbols.
\paragraph{Constants}
Constants are function symbols with an arity of zero.
The name of a constant can also be encompassed by double quotation marks, which is sometimes helpful when using constants with spaces. It has to be remarked that a constant with quotation marks is never equal to a constant without, so ``\texttt{a}'' is not equal to \texttt{a}.
$\mathcal{C}$ is the set of constants
\paragraph{Variables}
Variables have to start with a capitalized letter and can contain digits and letters.
The set of variables is denoted by $\mathcal{V}$.
The language over $\mathcal{P}$, $\mathcal{F}$ and $\mathcal{C}$ is denoted by $\mathcal{L(P,F,C)}$.
\paragraph{Terms}
The building blocks of the language are terms.
Each constant $c \in \mathcal{C}$ and each function symbol $f \in \mathcal{F}$ is a term.
Furthermore $f(t_1, \dots, t_n)$ is a term if $f/n \in \mathcal{F}$ and $t_i \mid i \in \{1 \dots n\}$ are terms.
Since a term is defined recursively, it can be nested.
A term with the first parameter being a constant $a$, the others being terms $f(b,X)$ and $g(f(X), Y))$ would be expressed this way.
\begin{verbatim}
 p(a,f(b,X), g(f(X),Y)).
\end{verbatim}
\paragraph{Atoms}
An atom over $\mathcal{L(P, F, C)}$ is of the form $p(t_1, \dots, t_n)$ where $p/n$ is a predicate
symbol with arity $n$, and $t_i \mid i \in \{1 \dots n\}$ are terms.
\paragraph{Rules}
A rule over the language $\mathcal{L(P,F,C)}$ has the form
$$
A_0 \mid \dots \mid A_l \leftarrow A_{l+1}, \dots, A_m, not\ A_{m+1}, \dots, not\ A_n.
$$
where $A_0$ to $A_n$ are atoms. \newline
A normal logic program $\Pi$ over $\mathcal{L(P,F,C)}$ is a finite set of rules over $\mathcal{L(P,F,C)}$.

@@

\subsection{How to write a logic program}
Writing a logic program to solve a problem is rather describing a problem than describing an algorithm to solve a problem.
Therefore the problem must be encoded as a set of rules.
Typically these are divided into a problem description and an instance of the problem.
The solution to a problem can then be read from the answer sets of the program.
The procedure of writing a logic program shall now be explained with a little example.
We want to write a problem encoding for the 3-colorability problem of graph coloring.
Given an arbitrary graph, no two connected edges may have the same color.
The most common technique is the ``generate \& test'' approach. Therefore, in the generator, rules are created that can derive a superset of the solution to the problem. In the tester all solutions that do not fit the constraints are omitted.
\newline
Usually you start with the non instance domains, in our example the colors we use.
\begin{verbatim}
 color(red). color(green). color(blue).
\end{verbatim}
After that, we start with the generator.
Therefore we have to color all the vertices, but we must give attention that each vertex has only one color.
We can do this by simply saying that a vertex is colored with one color $C$, and that the same vertex does not have any of the other colors $D$ or $E$.
\begin{verbatim}
 colored(V,C) :- not colored(V,D), not colored(V,E),
                 C != D, C != E, D != E,
                 color(E), color(C), color(D), vertex(V).
\end{verbatim}
As you can see, you have to give the domain of the variables in the rule.
The variable $V$ is used for vertices and the variables $C$, $D$ and $E$ are used for colors.
After every vertex now has a color we would get all possible solutions of all vertices colorfully painted.
We now need to restrict our solutions in the tester by saying that two vertices that are connected by an edge should not have the same color.
We can do this by using an integrity constraint.
An integrity constraint is a rule with an empty head, and if an assignment of variables fulfills the body, the assignment is not a solution of our problem.
\begin{verbatim}
 :- edge(V,U), colored(V,C), colored(U,C),
    vertex(V), vertex(U), color(C).
\end{verbatim}
Here we have the edge between two vertices $V$ and $U$, and if both are colored with the same color $C$ the assignment is not a solution to our problem.
\newline
We have finally written our problem description, all we need now is a problem instance. This step is fairly simple.
We just need to define several vertices and edges that do connect the vertices.
For the following simple graph
\[
 \xymatrix{1\ar@{-}[dr]\ar@{-}[d]\ar@{-}[r]&2\ar@{-}[dl]\\
           3&4}
\]

@@

\subsection{Features}
To ease the process writing logic programs some features in the input syntax are introduced and are now explained in detail.
For the most problems it is possible to write a much shorter version of the program with the extended syntax.
%Even a speedup in solving the instances of the problem can happen TODO ?
For a full syntax description in EBNF see appendix (\ref{appendix}).
\paragraph{Function Symbols}
Function symbols are fully supported with gringo2.0.
They can be finitely nested and also be used as domain predicates.
While in lParse function symbols are allowed in a very restricted way, no restrictions apply in gringo2.0.
The following program can not be grounded with lParse since $X$ is an unrestricted variable. With gringo2.0 the program is now groundable.
\begin{verbatim}
a(X) :- f(1,g(X)).
f(1,X) :- f(X).
f(g(2)).
\end{verbatim}
From the second rule the domain \texttt{f(1,g(2))} is generated.
This domain can be used for the first rule and results in the grounded rule
\begin{verbatim}
 a(2) :- f(1,g(2)).
\end{verbatim}

@@

\paragraph{Aggregates}
An aggregate is a set or a multiset of atoms and has a lower and an upper bound.
The general syntax is
\begin{verbatim}
 lower aggregate_name [atom1=w1, ..., atomN=wN] upper
\end{verbatim}
If squared brackets are used, the atom list is interpreted as a set, otherwise, for curly brackets, as a multiset of atoms.
The lower and upper bound is optional, and can be omitted.

@@

\subparagraph{Min}
The ``min'' aggregate is determined by the minimum of all weights of atoms that evaluate to true.
For the minimum aggregate a multiset with weighted atoms is used. It has no shorthand syntax.
\subparagraph{Max}
The ``max'' aggregate is similar to the minimum aggregate but considers the maximum of the weights.

@@

% fail if we have not found a better solution
 :- costs(C), C [ altserves(R,Dist) : dist(Dist) : restaurant(R,RK) = Dist ].

@@

%\section{Benchmarks}
%The benchmarks were run on an Athlon XP 1700+ with 1GB of RAM.
%TODO: Official (non internal wiki) link to the instances.

%\input{benchmark2.result}
\section{Future Work}
Further research has to be done on features in the input language. Additionally the support of incremental grounding will be improved. To speed up the grounding process research on a bottom up grounding procedure is done. To support an even less restricted input language the support for $\Pi$-restricted programs is planned in a future release of gringo. All these features will lead to more acceptance of ASP-systems as the programs become easier to write and the bottleneck of ASP, the grounding part, will be faster.
\newpage
\textbf{\appendixname}
\begin{appendix}
\section{EBNF Syntax}
\label{appendix}
\input{appendix}
\end{appendix}
